---
title: "CA_Wildfire_Exploration"
author: 
  - "William McDonough"
  - "AnnaMarie Y. Ho"
  - "Amira Cerimi"
date: "2024-05-07"
output:
  html_document: default
---

# Introduction

 By leveraging   the   comprehensive   wildfire   occurrence dataset from the 
 U.S. Department of Agriculture, this study seeks to uncover spatial hotspots 
 and temporal trends   in   California   and   investigate   potential
 correlations with temperature anomalies. Wildfires present significant economic
 and environmental challenges, with annual costs in the United States estimated 
 between 394 billion and 893 billion USD [1]. Given time-constraints, we limited
 our study to California after realizing the acreage burned there was very high
 compared to other states.
 

# Data and Transformations

[2] [2.3 Million US Wildfires (1992-2020) 6th Edition](https://www.kaggle.com/datasets/behroozsohrabi/us-wildfire-records-6th-edition)

```
Metadata (only including the data we used):

FIRE_YEAR Calendar year in which the fire was discovered or confirmed to exist.

DISCOVERY_DATE Date on which the fire was discovered or confirmed to exist.

DISCOVERY_DOY Day of year on which the fire was discovered or confirmed to exist.

DISCOVERY_TIME Time of day that the fire was discovered or confirmed to exist.

CONT_DATE Date on which the fire was declared contained or otherwise controlled (mm/dd/yyyy where mm=month, dd=day, and yyyy=year).

CONT_DOY Day of year on which the fire was declared contained or otherwise controlled.

CONT_TIME Time of day that the fire was declared contained or otherwise controlled (hhmm where hh=hour, mm=minutes).

FIRE_SIZE The estimate of acres within the final perimeter of the fire.

FIRE_SIZE_CLASS Code for fire size based on the number of acres within the final fire perimeter (A=greater than 0 but less than or equal to 0.25 acres, B=0.26-9.9 acres, C=10.0-99.9 acres, D=100-299 acres, E=300 to 999 acres, F=1000 to 4999 acres, and G=5000+ acres).

LATITUDE Latitude (NAD83) for point location of the fire (decimal degrees).

LONGITUDE Longitude (NAD83) for point location of the fire (decimal degrees).

STATE Two-letter alphabetic code for the state in which the fire burned (or originated), based on the nominal designation in the fire report (not from a spatial overlay).

COUNTY County, or equivalent, in which the fire burned (or originated), based on nominal designation in the fire report (not from a spatial overlay).
```

[3] [Climate at a Glance](https://www.ncei.noaa.gov/access/monitoring/climate-at-a-glance/statewide/time-series)
```
Metadata:
Date year and month formatted YYYYMM

Value average temperature in degrees farenheit

Anomaly difference in average and baseline (1901-2000 base period)
```


## Libraries
``` {r libraries_installation, message=FALSE, warning=FALSE, results='hide'}
# R packages
list.of.packages <- c('ggplot2', 
                      'dplyr', 
                      'ggiraphExtra', 
                      'sf', 
                      'maps', 
                      'mapproj',
                      'tigris',
                      'plotly',
                      'scales',
                      'zoo',
                      'corrplot',
                      'knitr',
                      'kableExtra',
                      'htmltools')

new.packages <- list.of.packages[!(list.of.packages %in%
                                     installed.packages()[,"Package"])]

if(length(new.packages)) {
  install.packages(new.packages, repo='http://cran.at.r-project.org',
                   dependencies=TRUE)
}

# Load all packages
lapply(list.of.packages, library, character.only = TRUE)
```


## Data Transformations

### Wildfire Data
After downloading the wildfire database, we need to read it into R and take the
`Fires` table

```{r get_fire_data}
# Install RSQLite if not already installed
if (!requireNamespace("RSQLite", quietly = TRUE)) {
  install.packages("RSQLite")
}

# Load the RSQLite package
library(RSQLite)

# Establish a connection to the SQLite database
db <- dbConnect(RSQLite::SQLite(), dbname = "../raw_data/fires_db.sqlite")

# Query the database, might take a second
query_result <- dbGetQuery(db, "SELECT * FROM Fires")

# The query_result is now a dataframe with the contents of 'Fires'
head(query_result)  # Verify this output with query_result.txt in Scripts

# Save the dataframe to disk as an R data file
saveRDS(query_result, file = "Data/fires_df.rds")

# Disconnect from the database
dbDisconnect(db)

# Get the columns of interest and rename them
fires <- query_result %>%
  select(FIRE_YEAR,
         STATE,
         COUNTY,
         DISCOVERY_DOY,
         CONT_DOY,
         FIRE_SIZE_CLASS,
         FIRE_SIZE,
         NWCG_CAUSE_CLASSIFICATION) %>%
  rename(year = FIRE_YEAR,
         state = STATE,
         county = COUNTY,
         day_discovered = DISCOVERY_DOY,
         day_contained = CONT_DOY,
         class = FIRE_SIZE_CLASS,
         size = FIRE_SIZE,
         cause = NWCG_CAUSE_CLASSIFICATION)

head(fires)
```

### Temperature Data
After downloading the CA temperatures csv we want to load it into a df, calculate
moving averages, and reformat dates for plotting. We are generating a df for both
monthly and annual temperatures

```{r get_temp_data}
# Load California temps (Base Period: 1901-2000)
ca_temps_df <- as.data.frame(read.csv("Data/ca_temps_monthly_raw.csv"))
head(ca_temps_df)

# Calculate moving averages
ca_temps_df$anomaly_mv_avg <- rollmean(ca_temps_df$Anomaly, 
                                   5, 
                                   fill = NA,
                                   align = "center")

ca_temps_df$temp_mv_avg <- rollmean(ca_temps_df$Value, 
                                       5, 
                                       fill = NA,
                                       align = "center")

# Convert these to the first day of the given month for accurate representation
ca_temps_df$Date <- as.Date(paste0(substr(ca_temps_df$Date, 1, 4), "-", 
                                   substr(ca_temps_df$Date, 5, 6), "-01"))
# Get month
ca_temps_df$month <- format(ca_temps_df$Date, "%m")

# Get year
ca_temps_df$year <- format(ca_temps_df$Date, "%Y")

# for some reason they are <chr>
ca_temps_df$year <- as.integer(ca_temps_df$year)
ca_temps_df$month <- as.integer(ca_temps_df$month)
str(ca_temps_df)

# GGplot needs custom labels or it just jumbles everything on the x axis
ca_temps_df$label <- ifelse(ca_temps_df$month == 01, as.character(ca_temps_df$year), "")

head(ca_temps_df, 13) # so you can see 2 labels

# need a df for monthly and annual 
ca_monthly_temps <- ca_temps_df %>%
  mutate(
    year_month = factor(paste(year, month, sep = "-"), levels = paste(year, month, sep = "-"))) %>%
  select(
    year, 
    month, 
    year_month, 
    Value, 
    Anomaly,
    temp_mv_avg,
    label) %>%
  rename(
    value = Value,
    anomaly = Anomaly) %>%
  arrange(year, month) 

ca_annual_temps <- ca_monthly_temps %>%
  select(year, value, label) %>%
  group_by(year) %>%
  summarise(annual_mean = mean(value, na.rm = TRUE),
            label = first(label),
            .groups = 'drop')

# Calculate moving average (didn't work in summarise())
ca_annual_temps <- ca_annual_temps %>%
  mutate(moving_avg = rollmean(annual_mean, k = 5, fill = NA, align = 'center'))

str(ca_annual_temps)
head(ca_annual_temps)
```
### Combined Data
Here we make simply combine the two datasets into a usable df and calculate some
moving averages for plotting:
```{r combined_data}
ca_fires <- fires %>%
  filter(state == "CA") %>%
  group_by(year) %>%
  summarise(number_of_fires = n()/100, # scaling for visual clarity
            natural_tot = sum(cause == "Natural")/100, # divide by 100 to
            human_tot = sum(cause == "Human")/100,    # match number_of_fires
            mean_acreage_burned = mean(size),
            pct_human = human_tot/number_of_fires*100)

head(ca_fires)

# join and get moving average
ca_combined_data <- left_join(ca_annual_temps, ca_fires, by = "year")
ca_combined_data <- ca_combined_data %>%
  mutate(fires_moving_avg = rollmean(number_of_fires, k = 5, fill = NA, align = 'center'),
         acres_moving_avg = rollmean(mean_acreage_burned, k = 5, fill = NA, align = 'center'),
         human_moving_avg = rollmean(human_tot, k = 5, fill = NA, align = 'center'),
         natural_moving_avg = rollmean(natural_tot, k = 5, fill = NA, align = 'center'),
         temp_moving_avg = rollmean(annual_mean, k = 5, fill = NA, align = 'center')) %>%
  rename(temp_annual_mean = annual_mean)

head(ca_combined_data)
```

This was used to get states, because `stateR` was not working.
```{r abbreviations}
state_abbreviations <- c(
  "Alabama" = "AL", "Alaska" = "AK", "Arizona" = "AZ", "Arkansas" = "AR", "California" = "CA",
  "Colorado" = "CO", "Connecticut" = "CT", "Delaware" = "DE", "Florida" = "FL", "Georgia" = "GA",
  "Hawaii" = "HI", "Idaho" = "ID", "Illinois" = "IL", "Indiana" = "IN", "Iowa" = "IA",
  "Kansas" = "KS", "Kentucky" = "KY", "Louisiana" = "LA", "Maine" = "ME", "Maryland" = "MD",
  "Massachusetts" = "MA", "Michigan" = "MI", "Minnesota" = "MN", "Mississippi" = "MS", "Missouri" = "MO",
  "Montana" = "MT", "Nebraska" = "NE", "Nevada" = "NV", "New Hampshire" = "NH", "New Jersey" = "NJ",
  "New Mexico" = "NM", "New York" = "NY", "North Carolina" = "NC", "North Dakota" = "ND",
  "Ohio" = "OH", "Oklahoma" = "OK", "Oregon" = "OR", "Pennsylvania" = "PA", "Rhode Island" = "RI",
  "South Carolina" = "SC", "South Dakota" = "SD", "Tennessee" = "TN", "Texas" = "TX",
  "Utah" = "UT", "Vermont" = "VT", "Virginia" = "VA", "Washington" = "WA", "West Virginia" = "WV",
  "Wisconsin" = "WI", "Wyoming" = "WY")
```

# Results

### Heatmap, Acreage Burned by State
Plotting the acreage burned by state illustrates why we chose CA for our first
exploration. Note that AK and HI are removed as their scale throws off the map
and eliminates any discernable differences in the continental states:
```{r heatmap_fires_by_state, message=FALSE, warning=FALSE, results='hide'}
# first get the acreage burned by state
state_total_acreage <- fires %>%
  mutate(state <- tolower(state)) %>%
  group_by(state) %>%
  summarise(tot_acreage = sum(size))

# generate a map
options(tigris_class = "sf")  # Ensure it returns sf objects
# cb = TRUE gives us a low-res version for quicker plotting
states_sf <- tigris::states(cb = TRUE)
states_sf <- states_sf %>%
  mutate(NAME = state_abbreviations[NAME])
states_sf <- states_sf %>%
  filter(!is.na(NAME))
# ensure you have the same identifiers
state_total_acreage$state <- trimws(state_total_acreage$state)
print(unique(states_sf$NAME))
print(unique(state_total_acreage$state))

# Merge using the common 'state' abbreviation column
choropleth_data <- merge(states_sf, 
                         state_total_acreage, 
                         by.x = "NAME", 
                         by.y = "state", 
                         all.x = TRUE)

choropleth_data <- choropleth_data %>%
  filter(!NAME %in% c("AK", "HI"))  # Exclude both Alaska and Hawaii due to scale
```

```{r plot_state_acreage}
ggplot_map <- ggplot(data = choropleth_data) +
  geom_sf(aes(fill = tot_acreage), 
          color = "white", 
          size = 0.1) +
  scale_fill_gradient(low = "yellow", 
                      high = "red", 
                      name = "Total Burned Acreage",
                      labels = scales::comma) +  # Adjust scale labels to standard numbering
  labs(title = "Total Burned Acreage by State (excluding Alaska & Hawaii)",
       subtitle = "Darker colors indicate more acreage burned. 1992-2020",
       x = NULL, # Remove axis titles
       y = NULL) +  # They are just coordinate labels
  theme_minimal() +
  theme(axis.text.x = element_blank(),  # Remove X axis text
        axis.text.y = element_blank(),  # Remove Y axis text
        axis.ticks = element_blank(),   # Remove axis ticks
        panel.grid.major = element_blank(),  # Remove major grid lines
        panel.grid.minor = element_blank(),  # Remove minor grid lines
        plot.background = element_blank()) +
  coord_sf(xlim = c(-125, -66), 
           ylim = c(24, 50), 
           expand = FALSE)

ggplot_map

```
### Linegraph, CA Temperature Anomalies

When looking at the anomalies, we see an observed increase in temperature anomaly
from around 2011. 
```{r linegraph_temp_anomaly, message=FALSE, warning=FALSE}
ca_anomalies_lineplot<- ggplot(ca_temps_df, aes(x = Date, y = Anomaly)) +
  geom_line(color = "black") +  # Adding a custom color to the line
  geom_point(aes(color = Anomaly), size = 2, alpha = 0.6) +  # Adding points
  geom_line(aes(y = anomaly_mv_avg), color = "darkred", size = 1.25) + # Anomaly moving Average
  scale_color_gradientn(colors = c("blue", "yellow", "red")) +  # Color gradient for points
  labs(
    title = "Temperature Anomalies in California",
    subtitle = "Monthly anomalies relative to the 1901-2000 average",
    x = "Year",
    y = "Anomaly (°F)",
    caption = "Data source: NOAA"
  ) +
  theme_minimal() +  # Minimalist theme
  theme(
    plot.background = element_rect(fill = "grey", color = "grey"),  # Set background to grey
    panel.background = element_rect(fill = "grey", color = "grey"),  # Match panel background
    text = element_text(color = "black"),  # Change text color to black for visibility
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12)
  ) +
  guides(color = guide_legend(title = "Temperature Anomaly"))

ca_anomalies_lineplot
```

### Barplots, California Temperatures

Here we plot monthly and annual temperatures to see any apparent trends:

Monthly doesn't offer much insight:
```{r barplot_monthly_temps, message=FALSE, warning=FALSE}
# for plot
overall_value_mean <- mean(ca_monthly_temps$value, na.rm = TRUE)

monthly_barplot <- ggplot(ca_monthly_temps, aes(x = year_month, y = value, fill = value)) +
  geom_col(width = 1) +  # width = 1 for no gaps
  geom_line(aes(y = temp_mv_avg, group = 1, color = "Moving Average"), size = 1) +
  geom_line(aes(y = overall_value_mean, group = 1, color = "Mean for Period"), size = 1) +
  labs(title = "Monthly Average Temperatures",
       subtitle = "California, 1992-2020",
       x = "Year",
       y = "Temperature (°F)",
       color = "Legend") +
  theme_minimal() +
  theme(plot.background = element_rect(fill = "grey", color = "grey"),  # Set background to grey
        panel.background = element_rect(fill = "grey", color = "grey"),  # Match panel background
        text = element_text(color = "black"),  # Change text color to black for visibility
        axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "right",
        plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12)) + # Rotate x-axis labels for clarity
  scale_x_discrete(labels = ca_monthly_temps$label) +
  coord_cartesian(ylim = c(38, 80)) + 
  scale_fill_gradient2(low = "yellow",
                       mid = "orange",
                       high = "red",
                       midpoint = overall_value_mean,
                       limits = c(min(ca_monthly_temps$value, na.rm = TRUE),
                                  max(ca_monthly_temps$value, na.rm = TRUE)),
                       name = "Temperature (°F)") +
  scale_color_manual(values = c("Moving Average" = "darkred", 
                                "Mean for Period" = "darkblue"))  

print(monthly_barplot)
```

However, annually we see some interesting trends:
```{r barplot_annual_temps, message=FALSE, warning=FALSE}
annual_barplot <- ggplot(ca_annual_temps, aes(x = year, y = annual_mean, fill = annual_mean)) +
  geom_col(width = 1) +
  geom_line(aes(y = moving_avg, group = 1, color = "Moving Average"), size = 1) +
  geom_line(aes(y = overall_value_mean, group = 1, color = "Mean for Period"), size = 1) +
  labs(title = "Annual Average Temperatures",
       subtitle = "California, 1992-2020",
       x = "Year",
       y = "Temperature (°F)",
       color = "Legend") +
  theme_minimal() +
  theme(plot.background = element_rect(fill = "grey", color = "grey"),  # Set background to grey
        panel.background = element_rect(fill = "grey", color = "grey"),  # Match panel background
        text = element_text(color = "black"),  # Change text color to black for visibility
        axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "right",
        plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12)) +
  coord_cartesian(ylim = c(55, 63)) + 
  scale_fill_gradient2(low = "yellow",
                       mid = "orange",
                       high = "red",
                       midpoint = overall_value_mean,
                       limits = c(min(ca_annual_temps$annual_mean, na.rm = TRUE),
                                  max(ca_annual_temps$annual_mean, na.rm = TRUE)),
                       name = "Temperature (°F)") +
  scale_color_manual(values = c("Moving Average" = "darkred", 
                                "Mean for Period" = "darkblue"))

print(annual_barplot)
```

It appears the temperature has been increasing from 2011 onward, coinciding with
the observed increase in temperature anomaly. The question is now whether this 
trend correlates to an increase in annual fires reported or annual acreage burned.

### Bar and Line Plot, California Wildfires and Temperatures

```{r bar_line_combined_data, message=FALSE, warning=FALSE}
combined_barplot <- ggplot(ca_combined_data, aes(x = year, y = number_of_fires, fill = temp_annual_mean)) +
  geom_col(width = 1) +
  geom_line(aes(y = human_moving_avg, group = 1, color = "Human-Caused"), size = 1) +
  geom_line(aes(y = natural_moving_avg, group = 1, color = "Natural-Caused"), size = 1) +
  geom_line(aes(y = fires_moving_avg, group = 1, color = "Number of Fires"), size = 1) +
  geom_line(aes(y = acres_moving_avg, group = 1, color = "Acreage"), size = 1) +
  labs(title = "Annual Wildfires and Temperatures",
       subtitle = "California, 1992-2020",
       x = "Year",
       y = "Number of Fires (in hundreds)",
       color = "Moving Averages") +
  theme_minimal() +
  theme(plot.background = element_rect(fill = "grey", color = "grey"),
        panel.background = element_rect(fill = "grey", color = "grey"),
        text = element_text(color = "black"),
        axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "right",
        plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12)) +
  # coord_cartesian(ylim = c(45, 175)) +
  scale_fill_gradient2(low = "yellow",
                       mid = "orange",
                       high = "red",
                       midpoint = mean(ca_combined_data$temp_annual_mean, na.rm = TRUE),
                       limits = c(min(ca_combined_data$temp_annual_mean, na.rm = TRUE),
                                  max(ca_combined_data$temp_annual_mean, na.rm = TRUE)),
                       name = "Mean Temperature (°F)") +
  scale_color_manual(values = c("Number of Fires" = "darkblue",
                                "Acreage" = "steelblue",
                                "Human-Caused" = "darkred",
                                "Natural-Caused" = "darkgreen")
                     )

print(combined_barplot)
```
Here we see an obvious sharp increase in the amount of acreage burned following 2010. 
Also note the overall decrease and apparent leveling of the percentage of human-
caused fires. 

We have observed an apparent increase in temperature coinciding with a sharp increase
in acres burned. Now we move on to statistical testing. How are these variables
related and is that relationship statistically meaningful?

### Statistics

```{r corplot}
# Correlation Testing

fires_pre_cor <- ca_combined_data %>%
  select(number_of_fires,
         temp_annual_mean,
         mean_acreage_burned,
         pct_human) %>%
  rename(annual_fires = number_of_fires,
         annual_temp = temp_annual_mean,
         annual_acreage_burned = mean_acreage_burned,
         percent_human_caused = pct_human)

fires_cor <- cor(fires_pre_cor)
corrplot(fires_cor, method = "number")

col <- colorRampPalette(c("#BB4444",
                          "#EE9988",
                          "#FFFFFF",
                          "#77AADD",
                          "#4477AA"))

corrplot(fires_cor,
         method="color",
         col=col(200),
         type="lower",
         order="hclust",
         addCoef.col="black",
         tl.col="black",
         tl.srt=45,
         diag=F)

cor_test_result <- cor.test(fires_pre_cor$annual_temp, fires_pre_cor$annual_acreage_burned)

cor_test_df <- data.frame(
  Estimate = cor_test_result$estimate,
  Statistic = cor_test_result$statistic,
  DF = cor_test_result$parameter,
  P.value = cor_test_result$p.value,
  Confidence.Interval = paste(round(cor_test_result$conf.int[1], 3), "to", round(cor_test_result$conf.int[2], 3))
)

# Use kable() for table
table_output <- kable(cor_test_df, format = "html", caption = "Temp & Acreage Correlation") %>%
  kable_styling(full_width = FALSE, position = "left", bootstrap_options = c("striped", "hover", "condensed", "bordered")) %>%
  column_spec(1, bold = TRUE, border_right = TRUE) %>%
  row_spec(0, bold = TRUE, color = "black", background = "lightgrey")

table_output
```
Notes:
annual fires is not meaningfully correlated to how many are human-caused. 
annual temperature has a negative correlation with percent human-caused.
annual acreage burned has a negative correlation with percent human-caused and 
positive correlations with mean annual fires and mean annual temperature.

Next we will explore how significant the acreage correlations are:
```{r linear_regression}
# Linear Regression acreage v year, temp, and pct human

model <- lm(mean_acreage_burned ~ year + 
              temp_annual_mean + 
              pct_human, 
            data = ca_combined_data)
par(mfrow = c(2, 2))
plot(model)
summary(model)
```
Note point 29 (year 2020) seems to have a high residual and a high leverage, 
indicating it is having substantial influence on the model. We can explore this
further to decide if we should remove the point:
```{r point_29_eval}
# point 29 seems to be influencing the regression model a lot
point_29 <- ca_combined_data[29, ]
print(point_29)

# Calculate leverage
influence_measures <- influence.measures(model)
print(influence_measures)

hat_values <- hatvalues(model)
print(hat_values[29]) # a measure of leverage (avg 0.276 for this data)

# Cook's distance
# This was returning NULL?
# print(influence_measures$cooks.distance[29])

# Manual Cook's
n = length(residuals(model))
p = length(coefficients(model))
h_29 = hat_values[29]
r_29 = residuals(model)[29]
cooks_d_29 = (r_29^2 / (p * mean((residuals(model))^2))) * (h_29 / (1 - h_29)^2)
print(cooks_d_29) # 1.29 means we can remove this (much higher than 4/n or 1)

# DFBETAS - to see the influence on each coefficient
# DFBETA is |2/sqrt(n)| so 2/sqrt(29) ~ 0.372
print(dfbetas(model)[29, ]) # intercept 0.778, higher than 0.372
```
Given the cook's value of 1.29 and all DFBETAS being outside the range, we exclude
point 29 for the model:
```{r regression_without_29}
# try without 29
model_excl_29 <- lm(mean_acreage_burned ~ year + temp_annual_mean + pct_human, data = ca_combined_data[-29, ])
summary(model_excl_29)
```

#### Key Points about the Statistics

The intercept has a large negative value and is not statistically significant 
(p > 0.05). The magnitude has decreased compared to the model with observation 
29 included, suggesting that the exclusion of this outlier has a leveling effect 
on the intercept

year: The coefficient for year is slightly positive but remains statistically 
insignificant. This suggests that there is not a clear trend over the years that 
is consistent across the data, at least not one strong enough to be statistically
significant without the outlier.

temp_annual_mean: The coefficient here is statistically significant (p < 0.01) 
and indicates a positive relationship. This suggests that temperature has a 
meaningful and positive impact on mean acreage burned, which becomes more 
pronounced without the outlier. This coefficient has become even more statistically
significant than in the full model, suggesting a clearer effect of temperature
on acreage burned without the distortion caused by the outlier.

pct_human: This coefficient remains statistically insignificant and indicates a 
minimal impact of the percentage of human-caused fires on the burned acreage in 
this model setup. The effect size and significance level have not notably changed, 
suggesting that the outlier did not heavily influence this relationship.

F-statistic: Shows that the overall model is statistically significant with a 
lower p-value than in the full model, suggesting that the model fits the data 
better without the outlier.

Adjusted R-squared: Increased significantly, reinforcing that the model explains 
more variability in a way that is adjusted for the number of predictors used,
which is more appropriate when comparing models with different numbers of predictors.

Residual standard error: The RSE has decreased from 63.57 in the full model to 
33.56 in this model, suggesting better prediction accuracy without the outlier.

To summarise concisely, with or without the outlier point 29 we observe a strong 
statistically significant positive correlation between the mean temperature in a year
and the amount of acreage burned by wildfires in California.

We do not, however, see a meaningful relationship with the percentage of human-caused
fires and the acreage burned.

# References
--------------------------------------------------------------------------------
[1] Joint Economic Committee Democrats. (2023, October). Climate-Exacerbated 
Wildfires Cost the U.S. Between $394 to $893 Billion Each Year in Economic Costs
and Damages. https://www.jec.senate.gov/public/index.cfm/democrats/2023/10/climate-
exacerbated-wildfires-cost-the-u-s-between-394-to-893-billion-each-year-in-economic-
costs-and-damages

[2] Short, Karen C. 2022. Spatial wildfire occurrence data for the United States,   
1992-2020 [FPA_FOD_20221014]. 6th Edition. Fort Collins, CO: Forest Service Research   
Data Archive. https://doi.org/10.2737/RDS-2013-0009.6

[3] NOAA National Centers for Environmental information, Climate at a  Glance:   
Statewide Time Series, published May 2024. https://www.ncei.noaa.gov/access/monitoring/
climate-at-a-glance/statewide/time-series   
